---
title: "Analysis"
format: html
editor: visual
---

```{r}
library(tidyverse)

df <- read_csv("Students Social Media Addiction.csv")

df <- df|>
  select(Student_ID, Academic_Level, Gender, Avg_Daily_Usage_Hours, Sleep_Hours_Per_Night, Mental_Health_Score, Addicted_Score) |>
  mutate(Academic_Level = as.factor(Academic_Level),
         Gender = as.factor(Gender)
         )

head(df)
```

## Research Question 1:

Do daily usage hours, sleep duration, mental health score, and addiction score jointly differ across academic levels?

### Hypothesis

$$
H_o: \vec{\mu_{1}} = \vec{\mu_{2}} = \vec{\mu_{3}} \text{ vs. } H_A: \text{At least two } \vec{\mu_{j}} \text{ differ for } j = {1,2,3}
$$

$$
\mu_j =
\begin{cases}
\text{True mean Avg\_Daily\_Usage\_Hours for academic level } j \\
\text{True mean Sleep\_Hours\_Per\_Night for academic level } j \\
\text{True mean Mental\_Health\_Score for academic level } j \\
\text{True mean Addicted\_Score for academic level } j
\end{cases}
$$

$$
j \in \{\text{High School}, \text{Undergraduate}, \text{Graduate}\}
$$

### MANOVA Test

```{r}
manova_fit <- manova(
  cbind(
    Avg_Daily_Usage_Hours,
    Sleep_Hours_Per_Night,
    Mental_Health_Score,
    Addicted_Score
  ) ~ Academic_Level,
  data = df
)

summary(manova_fit, test = "Wilks")
summary(manova_fit, test = "Pillai")
summary(manova_fit, test = "Hotelling-Lawley")
summary(manova_fit, test = "Roy")
```

Based on the MANOVA results (e.g., Wilks’ Lambda / Pillai’s Trace, etc.), the approximate F values corresponding to the test statistics are all large resulting in small p-values (p \< 0.05), so we reject $H_o$ Thus, Academic Level has a significant joint effect on the four response variables (Average Daily Usage Hours, Sleep Hours Per Night, Mental Health Score, and Addicted Score).

### Assumptions

```{r}
check.mvnorm.plot <- function(data) {
  data <- as.matrix(data)
  
  d2 <- numeric(nrow(data))
  
  xbar <- colMeans(data)
  S <- cov(data)
  
  for (i in 1:nrow(data)) {
    diff <- as.matrix(data[i, ] - xbar)
    d2[i] <- t(diff) %*% solve(S) %*% diff
  }
  
  p <- ((1:nrow(data)) - 0.5) / nrow(data)
  
  plot(
    qchisq(p, df = ncol(data)),
    sort(d2),
    ylab = "D^2 Values",
    xlab = "Theoretical Chi-Square Quantiles",
    main = "Chi-Square Probability Plot"
  )
  
  abline(0, 1)
}

highschool <- df |>
  filter(Academic_Level == "High School") |>
  select(Avg_Daily_Usage_Hours, Sleep_Hours_Per_Night, Mental_Health_Score, Addicted_Score)

undergrad <- df |>
  filter(Academic_Level == "Undergraduate") |>
  select(Avg_Daily_Usage_Hours, Sleep_Hours_Per_Night, Mental_Health_Score, Addicted_Score)

graduate <- df |>
  filter(Academic_Level == "Graduate") |>
  select(Avg_Daily_Usage_Hours, Sleep_Hours_Per_Night, Mental_Health_Score, Addicted_Score)

check.mvnorm.plot(highschool)
check.mvnorm.plot(undergrad)
check.mvnorm.plot(graduate)
```

We used Chi-square probability plots to assess multivariate normality within each academic level. The plots were fairly linear in the center, but there were noticeable deviations in the upper tails for two groups, suggesting some departure from multivariate normality due to heavy tails or possible outliers. Thus, the assumption appears to be approximately met, so this is good enough to go along with our analysis but we should be cautious of this.

```{r}
Box.M <- function(samp.covs, n) {
  k <- length(n)
  p <- nrow(samp.covs[[1]])
  df <- n - 1
  
  S.dets <- numeric(k)
  Sp <- matrix(0, nrow = p, ncol = p)
  
  for (i in 1:k) {
    S.dets[i] <- det(samp.covs[[i]])
    Sp <- Sp + df[i] * samp.covs[[i]]
  }
  
  Sp <- Sp / sum(df)
  det.Sp <- det(Sp)
  
  # compute log(M) directly to avoid overflow/underflow --- 
  #determinants were to small for original function so had to modify to this for it to work
  logM <- sum((df / 2) * log(S.dets)) - (sum(df) / 2) * log(det.Sp)
  
  c <- (sum(1 / df) - 1 / sum(df)) *
    ((2 * p^2 + 3 * p - 1) / (6 * (p + 1) * (k - 1)))
  
  u <- -2 * (1 - c) * logM
  u.df <- 0.5 * (k - 1) * p * (p + 1)
  p.value <- 1 - pchisq(u, u.df)
  
  return(list(u = u, p.value.u = p.value, chi.df = u.df))
}

samp.covs <- list(
  cov(highschool),
  cov(undergrad),
  cov(graduate)
)

n <- c(
  nrow(highschool),
  nrow(undergrad),
  nrow(graduate)
)

Box.M(samp.covs, n)
```

Box’s M test was conducted to evaluate the MANOVA assumption of equal covariance matrices across academic levels. The result was significant (u = 161.937, df = 20, p-value \< 0.001), suggesting that the covariance matrices are not the same for all academic level groups. Therefore, this assumption is violated, and the MANOVA results should be interpreted with some caution.

### Univariate ANOVA follow ups

```{r}
summary.aov(manova_fit)
```

Thanks to rejecting the null hypothesis for the MANOVA test we can perform follow-up univariate ANOVAs which showed that academic level had a significant effect on all four response variables.

We did not meet our assumptions of equal population covariances for the MANOVA tests so we do need to keep that in mind and treat these results with caution.

1.  For average daily usage hours, there was a significant difference across academic levels (F(2, 702) = 6.2655, p = 0.002009), indicating that at least one academic level has a different mean daily social media usage time than another.

2.  For sleep hours per night, there was also a significant difference across academic levels (F(2, 702) = 26.593, p = 7.368e-12), suggesting that average sleep duration is not the same for all academic groups.

3.  Mental health score likewise differed significantly by academic level (F(2, 702) = 17.72, p = 3.108e-08), showing that at least one group has a different mean mental health score.

4.  Finally, addiction score also showed a significant difference across academic levels (F(2, 702) = 17.114, p = 5.534e-08), indicating that average social media addiction score varies among the academic levels.

Overall, these results suggest that academic level is associated with differences in daily social media usage, sleep duration, mental health, and addiction score.

## Research Question 2:

What underlying patterns exist among social media usage, sleep, mental health, and addiction, and do these variables group students into meaningful clusters?

### PCA

We use PCA to reduce the dimensionality of the data while still keeping most of the information. Since the four variables are likely related, PCA helps summarize them into a smaller number of components that capture the main patterns in the data.

```{r}
princomp.test <- function(pr.out){
lambda <- pr.out$sdev^2
n <- pr.out$n.obs
p <- length(lambda)
mult <- n - ((2*p + 11)/6)
u <- matrix(0,p-1)
df <- matrix(0,p-1)
p.val <- matrix(0,p-1)
for(k in 2:p){
l.bar <- mean(lambda[(p-k+1):p])
u[k-1] <- mult * (k*log(l.bar) - sum(log(lambda[(p-k+1):p])))
df[k-1] <- .5*(k-1)*(k+2)
p.val[k-1] <- 1 - pchisq(u[k-1], df[k-1])
}
out <- cbind(lambda[1:(p-1)], p:2, rev(u), rev(df), rev(p.val))
dimnames(out)[[2]] <- c("Eigenvalue", "k", "u","df","p-value")
return(out)
}
```

```{r}
pca_data <- df[, c(
  "Avg_Daily_Usage_Hours",
  "Sleep_Hours_Per_Night",
  "Mental_Health_Score",
  "Addicted_Score"
)]

pca_scaled = scale(pca_data)
```

```{r}
pca_fit <- princomp(pca_scaled, cor = FALSE)
summary(pca_fit)
loadings(pca_fit)
```

```{r}
princomp.test(pca_fit)
```

```{r}
plot(pca_fit, type = "lines", main = "Scree Plot")
```

#### Results and interpretation

The first principal component explains about 85.6% of the total variation, which means that most of the information in the data can be represented along a single dimension. The scree plot also shows a clear drop after the first component, supporting this conclusion.

Looking at the ladings, the first principal component has positive contributions from average daily usage and addiction score, and negative contributions from sleep hours and mental health score. This suggests that the main pattern in the data reflects a contrast between higher usage and addiction versus lower sleep and poorer mental health.

The second principal component explains much less variation, about 8.5%, and captures smaller differences between variables, mainly involving sleep and mental health. Since it contributes relatively little, the primary interpretation focuses on the first component.

The PCA significance test shows very small p-values, indicating that the leading components are statistically meaningful and that reducing the data to fewer dimensions is appropriate.

### Hierarchical Clustering

Cluster analysis is used to group observations based on similarity across multiple variables. In this case, it helps identify whether there are distinct types of students with similar patterns of usage, sleep, mental health, and addiction.

We use Euclidean distance on standardized variables so that each variable contributes equally. Complete linkage is applied because it forms more compact and interpretable clusters by considering the maximum distance between observations in different clusters.

```{r}
cluster_vars <- df[, c(
  "Avg_Daily_Usage_Hours",
  "Sleep_Hours_Per_Night",
  "Mental_Health_Score",
  "Addicted_Score"
)]
cluster_scaled <- scale(cluster_vars)
```

```{r}
D <- round(dist(cluster_scaled), 2)
D
```

```{r}
dist_mat <- round(dist(cluster_scaled[1:5, ], diag = TRUE, upper = TRUE), 2)
dist_mat
```

```{r}
hc_fit <- hclust(dist(cluster_scaled), method = "complete")
hc_fit
```

```{r}
plot(hc_fit,
     labels = FALSE,
     ylab = "Maximum Distance Between Clusters",
     main = "Dendrogram for Complete Linkage Clustering")

rect.hclust(hc_fit, k = 3, border = "red")
```

```{r}
plot(rev(hc_fit$height),
     type = "b",
     xlab = "Merge Step",
     ylab = "Merge Distance",
     main = "Merge Heights Plot")

abline(v = 3, col = "red", lty = 2)
```

```{r}
cluster_membership <- cutree(hc_fit, k = 3)
table(cluster_membership)
```


The dendrogram suggests that dividing the data into three clusters provides a reasonable grouping. The clusters are also fairly balanced in size, with 217, 284, and 204 observations.

Cluster 2 represents a lower-risk group, with lower usage and addiction scores and higher sleep and mental health values. Cluster 3 represents a higher-risk group, with higher usage and addiction along with lower sleep and poorer mental health. Cluster 1 falls in between and reflects a more moderate profile across all variables.

Overall, the clusters show a clear pattern that aligns with the main direction identified in the PCA. Students tend to fall along a spectrum from lower-risk to higher-risk behavioral profiles.
